### 2013-10-01 Publications/2013-ISBI14/02-introduction.tex
Chus' changes

\Gls*{fmb} methods supposed the first approach to retrospective
correction \cite{jezzard_correction_1995}. They rely on one
extra acquisition (\emph{field mapping}), that probes the 
inhomogeneity of the B0 field.
A second theory-based breed of methodologies acquire a 
map of the \acrlong*{psf} of the \gls*{epi} readouts to correct 
for the artifact \cite{robson_measurement_1997}. 
Next generation of methodologies \cite{cordes_geometric_2000,
chiou_simple_2000}, make use of the acquisition of 
extra \gls*{epi} volumes 
with specific differences that enable correction, for instance
the opposed direction of the phase encoding gradients. We will
refer to these methods as \gls*{reb}.
Finally, the last family of methodologies acquire an 
extra T2-weighted image, and use its anatomical correctness 
as reference to find the deformation map through nonlinear 
registration \cite{kybic_unwarping_2000,studholme_accurate_2000}.
\gls*{t2b} methods usually map
the T2 image to the \emph{baseline} volume or \textit{b0}
of \gls*{dmri}. The choice of T2 is due to the strong
similarity of intensity distribution with respect the \textit{b0}.
More recent works report extensions or combined
approaches of existing techniques
\cite{andersson_how_2003,zaitsev_point_2004,%
holland_efficient_2010,andersson_comprehensive_2012}.

Even though aforementioned techniques for distortion correction
have been studied \cite{zeng_image_2002,wu_comparison_2008},
the lack of a gold-standard limits the possible benchmarking 
strategies. Recently, \cite{irfanoglu_effects_2012} raised the question
of distortion-derived impacts in tractography.
In this work, we propose an evaluation framework
using a digital phantom designed for connectivity assessment.
This framework enabled us to compare several correction 
techniques and characterize their geometrical accuracy 
and the dropout compensation. Finally, we report their 
impact on subsequent tractography and resulting connectivity
matrices.

>>>>

\Gls*{fmb} methods \cite{jezzard_correction_1995} are
the initial approach to retrospective corrections. 
They rely on one extra acquisition (\emph{field mapping}),
that probes the inhomogeneity of the B0 field.
A second theory-based breed of methodologies acquire a 
map of the \acrlong*{psf} of the \gls*{epi} readouts to correct 
for the artifact \cite{robson_measurement_1997}. 
Next generation of methodologies \cite{cordes_geometric_2000,
chiou_simple_2000}, make use of the acquisition of 
extra \gls*{epi} volumes 
with specific differences that enable correction, for instance
the opposed direction of the phase encoding gradients. We will
refer to these methods as \gls*{reb}.
Finally, the last family of methodologies acquire an 
extra T2-weighted image, and use its anatomical correctness 
as reference to find the deformation map through nonlinear 
registration \cite{kybic_unwarping_2000,studholme_accurate_2000}.
\gls*{t2b} methods usually map
the T2 image to the \emph{baseline} volume or \textit{b0}
of \gls*{dmri}. The choice of T2 is due to the strong
similarity of intensity distribution with respect the \textit{b0}.
More recent works report extensions or combined
approaches of existing techniques
\cite{andersson_how_2003,zaitsev_point_2004,%
holland_efficient_2010,andersson_comprehensive_2012}.

Even though aforementioned techniques for distortion correction
have been studied \cite{zeng_image_2002,wu_comparison_2008},
the lack of a gold-standard limits benchmarking 
strategies. Recently, Irfanoglu et al.
\cite{irfanoglu_effects_2012} raised the question
of distortion-derived impacts in tractography.
In this work, we propose an evaluation framework
using a digital phantom designed for connectivity assessment.
This framework enabled us to compare several correction 
techniques and characterize their geometrical accuracy 
and the dropout compensation. Finally, we report their 
impact on subsequent tractography and resulting connectivity
matrices.

### 2013-10-01 Publications/2013-ISBI14/03-methods.tex
Chus' changes

\subsection{Digital Phantom}
Based on the fiber geometries of the digital phantom 
created for the High angular resolution diffusion imaging 
reconstruction Challenge held in ISBI 2013 
(San Francisco, US), we simulated high resolution 
(0.5mm isotropic) T1 (TE/TR=10/1500ms) and T2 
(TE/TR=90/5000ms) images, as well as a \gls*{dmri}
(1.0mm isotropic, $b$=1200, 32 evenly-distributed 
directions, 1 \textit{b0} image) image.
The phantom includes \gls*{wm} fiber bundles, 
\gls*{csf} and \gls*{gm}. Diffusion is modeled by a 
restricted and a hindered compartment, similar to
\cite{assaf_composite_2005}.

@@

Once a realistic \gls*{vsm} was available, we generated
the corresponding distorted \glspl*{dwi}, in two opposed
phase-gradient encoding directions. The second simulated
``acquisition'' of the same phantom was necessary 
for evaluating the second family of correction methods.

In this simple manner, we generated a full gold-standard
containing realistic T1 and T2
at high resolution, \glspl*{dwi} acquired in two
different phase encoding directions, and a ground-truth
\gls*{dwi} data, which is not available for real datasets.

@@

Three out of four methods presented in \autoref{sec:intro}
were tested on the evaluation framework.

\paragraph*{\Gls*{fmb}} In this case, the pipeline essentially
the same as the one used for generating the distortion, but
applying the \gls*{vsm} in the opposite direction.

\paragraph*{\Gls*{reb}} We made use of the implementation found
in \emph{FSL} (called \texttt{topup}) to correct using the \textit{b0} of a
reverse-encoding simulation of the \glspl*{dwi}. In this second
case, a \gls*{vsm} is inferred from the differences between the 
two corresponding \textit{b0}.

\paragraph*{\Gls*{t2b}} For the last surveyed method,
we fine-tuned \emph{ANTs} \cite{avants_ants:_2013},
in order to register \textit{b0} to T2. To this end, we used a
multi-resolution scheme with 3 levels of subsampling and smoothing,
mutual information metric, and symmetric diffeomorphic transform 
(\texttt{SyN}). Several configurations of kernel widths for the 
regularization smoothers were tested, and finally selected 
0.5/1.0 voxels (gradient/deformation fields, respectively) 
for its best result. Additionally, undistorted images are 
corrected for dropout using the determinant
of the Jacobian of the deformation field.

@@

In order to enable the analysis of the influence of
distortions on the final connectivity analyses,
a intermediate step was also implemented to
provide with deterministic tractography,
using the \emph{Diffusion Toolkit} \cite{wang_diffusion_2007}.
The original phantom, one distorted version, and
the corrected instances are fed to this module.
Additionally, the original tissue probability maps
are also distorted and corrected, for computing overlap
indices.
Each tractography was restricted by its corresponding 
\gls*{wm} mask (e.g. the distorted and corrected with
fieldmap for the tractography that followed this method). 
In all cases, we used 10 random seeds at each seeding
voxel and the seeding voxels were taken from the 
ground-truth, mimicking the procedure with real data
(in which this regions are typically projected from the
anatomically correct T1).

Finally, the proposed workflows are connected to performance
assessment modules. We evaluated three characteristics
of the correction methods. 
First, we assessed the geometrical correctness
and reported overlap indices of three tissue
probability maps (namely \gls*{csf}, \gls*{wm},
and \gls*{gm}),
weighting the average by the volumes of labels.
Second, to evaluate the quality of the actual 
signal dropout correction, we studied the 
signal similarity frame by frame computing the $\ell_1$-norm
correlation index. We report this score
on the \textit{b0} and the average of the remaining 32 
\glspl*{dwi}. Third, we studied the impact on the
connectivity matrices reporting the number of
false positives (inexistent connections in the
gold-standard) and false negatives (or connections
lost).

>>>>

\subsection{Digital Phantom}
Based on the fiber geometries of the digital phantom 
created for the \emph{High angular resolution diffusion imaging 
reconstruction Challenge} held in ISBI 2013 
(San Francisco, US), we simulated high resolution 
(0.5mm isotropic) T1 (TE/TR=10/1500ms) and T2 
(TE/TR=90/5000ms) images, as well as a \gls*{dmri}
(1.0mm isotropic, $b$=1200, 32 evenly-distributed 
directions, 1 \textit{b0} volume) image.
The phantom includes \gls*{wm} fiber bundles, 
\gls*{csf} and \gls*{gm}. Diffusion is modeled by a 
restricted and a hindered compartment, similar to
\cite{assaf_composite_2005}.

@@

Once a realistic \gls*{vsm} was available, we generated
the corresponding distorted \glspl*{dwi}, in two opposed
phase-gradient encoding directions. The second simulated
``acquisition'' of the same phantom was necessary 
for evaluating \gls*{reb} methods.

Therefore, following the described strategy we generated 
a full gold-standard containing realistic T1 and T2
at high resolution, \glspl*{dwi} acquired in two
different phase encoding directions, and a ground-truth
\gls*{dwi} data, which is not available for real datasets.

@@

Three out of four methods presented in \autoref{sec:intro}
were tested on the evaluation framework. \Gls*{fmb} correction
is essentially the same as the one used for generating the 
distortion, using the synthetic fieldmap to infer a new \gls*{vsm}.
For practical reasons (distortion on the mask, signal dropout), the
method did not achieve a perfect geometrical correction as might be
expected if the process was exactly inverse. For \Gls*{reb} correction,
we made use of the implementation found in \emph{FSL} (\texttt{topup})
that demands for the \textit{b0} of the reverse-encoding simulation.
In this second case, the \gls*{vsm} is inferred from the differences
between the two corresponding \textit{b0}.

Finally, for evaluating \Gls*{t2b},
we fine-tuned \emph{ANTs} \cite{avants_ants:_2013},
in order to register \textit{b0} to T2. To this end, we used a
multi-resolution scheme with 3 levels of subsampling and smoothing,
mutual information metric, and symmetric diffeomorphic transform 
(\texttt{SyN}). Several configurations of kernel widths for the 
regularization smoothers were tested, and finally selected 
0.5/1.0 voxels (gradient/deformation fields, respectively) 
for its best result. Additionally, undistorted images are 
corrected for dropout using the determinant
of the Jacobian of the deformation field.

@@

In order to enable the analysis of the influence of
distortions on the final connectivity outcome,
an intermediate step was also implemented to
produce deterministic tractography,
using the \emph{Diffusion Toolkit} \cite{wang_diffusion_2007}.
The original phantom, one distorted version, and
the corrected instances are fed to this module.
Additionally, the original tissue probability maps
are also distorted and corrected, for computing overlap
indices.
Each tractography was restricted to its corresponding 
\gls*{wm} mask, after distortion and correction when
applied. For all the cases, we used 10 random seeds at each seeding
voxel and the seeding voxels were taken from the 
ground-truth, mimicking the usual procedure on real data
(where regions are typically mapped from the
anatomically correct T1).

The evaluation framework is completed by automated 
assessment modules. We evaluated three characteristics
of the correction methods. 
Firstly, we assessed the geometrical correctness
reporting overlap indices of three tissue
probability maps (namely \gls*{csf}, \gls*{wm},
and \gls*{gm}), weighting the average by tissue
volumes.
Secondly, to evaluate the quality of the actual 
signal dropout correction, we studied the 
signal similarity frame by frame computing the $\ell_1$-norm
correlation index. We report this score
on the \textit{b0} and the average of the remaining 32 
\glspl*{dwi}. Thirdly, we studied the impact on the
connectivity matrices reporting the number of
false positives (inexistent connections in the
gold-standard) and false negatives (or connections
lost).

### 2013-10-01 Publications/2013-ISBI14/04-results.tex
Last changes
Chus' changes

\section{RESULTS AND DISCUSSION}
\label{sec:results}
The results of the proposed experiments are summarized
in \autoref{fig:results} and \autoref{table:results}.
The remaining of this section provides extended descriptions
and interpretation of results.

@@

\begin{figure*}[tpb]
   \centering
   \includegraphics[width=0.9\textwidth]{Fig02-Results}
   \caption{The three major stages of the evaluation framework,
   along with the associated visual results. First row represents
   a coronal section of the \textit{b0} volume. In second row, the outcome
   of tractography, showing only tracks that connect two different
   network nodes. Third row shows the associated connectivity matrix. }
   \label{fig:results}
\end{figure*}

@@

A second workflow evaluated the different correction 
methodologies to show the similarity of the recovered
signal with respect to the original (undistorted).
In this second study, \gls*{reb} performed significantly
better than the other two methods, as reported
in \autoref{table:results}. \gls*{reb} scored a $91.0\%$
similarity index for the \textit{b0} volume and an average $97.65\%$
for the remaining 31 \glspl*{dwi}. Again, the second
qualified was \gls*{fmb}, which achieved very close results
for the \glspl*{dwi} ($96.26\%$) but not as good for the \textit{b0}
volume. Visual inspection of the recovered data confirm the 
results presented in the Table (\autoref{fig:results}, first
row). Finally, \gls*{t2b} confirmed that its correction
can not be compared, even when using the intensity compensation
based on the determinant of the deformation field.

@@

For the shake of completeness, we also document the outcome
of tractography after each correction method. After distortion,
we obtained 1750 tracks and a characteristic length of 
$40.64\pm8.12$mm.
\gls*{fmb} produced 1900 tracks/$40.53\pm8.07$mm., \gls*{reb}
obtained 2200 tracks/$41.15\pm7.60$mm., and 2500 
tracks/$40.60\pm7.81$mm. for \gls*{t2b}. Although track counting has
been used as a measurement of success and as a weight of connectivity 
strength, the number of tracks produced by tractography strongly rely
on the seeding performed and the smoothing caused by interpolations 
within correction methodologies.

@@

Even though all the surveyed methods produced visually 
sound results, our study suggested that \gls*{reb} is the 
most accurate methodology for correction of susceptibility-induced
distortions in \gls*{dwi}. Moreover, \gls*{reb} also achieved
the best results in terms of connectivity, being able to recover
one of two lost connections after distortion. \Gls*{fmb} achieved 
very positive scores. Finally, the \gls*{t2b} method did not achieve
the necessary high-standards to recommend its use. We understand
that specific methods with anisotropic regularization that completely
restrict deformations to the phase-encoding axis would perform
significantly better as compares to the standard methods in this study.

One relevant limitation of this work is that the \gls*{fmb} is
used both in distortion creation and correction, what could
bias results. Nonetheless, \gls*{reb} performed better that \gls*{fmb}.
A second limitation is found on the impossibility of integrating the
point spread function-based methods for two reasons.
First, the unmet need of a point spread function map consistent
with the synthetic phase-difference map. Second, to our knowledge,
there are no implementations of the method readily available.
In practice, this second limitation is narrowed down by the
fact that typical acquisition protocols usually include one or more
datasets among field mapping, reversed-encoding \textit{b0} or T2,
but they do not include the point-spread function mapping.

>>>>

\section{RESULTS}
\label{sec:results}
The results of the proposed experiments are summarized
in \autoref{fig:results} and \autoref{table:results}.
The remaining of this section provides extended descriptions
and interpretation of results.

@@

\begin{figure*}[tpb]
   \centering
   \includegraphics[width=0.9\textwidth]{Fig02-Results}
   \caption{Visual comparison of correction methods results. 
   First row represents
   a coronal section of the \textit{b0} volume. In second row, the outcome
   of tractography, showing only tracks that connect two different
   network nodes. Third row shows the associated connectivity matrix. }
   \label{fig:results}
\end{figure*}

@@

A second workflow investigated the similarity of the recovered
signal with respect to the original (undistorted).
In this second study, \gls*{reb} performed significantly
better than the other two methods, as reported
in \autoref{table:results}. \gls*{reb} scored a $91.0\%$
similarity index for the \textit{b0} volume and an average $97.65\%$
for the remaining 31 \glspl*{dwi}. Again, the second
qualified was \gls*{fmb}, which achieved very close results
for the \glspl*{dwi} ($96.26\%$) but not as good for the \textit{b0}
volume. Visual inspection of the recovered data confirm the 
results presented in the Table (\autoref{fig:results}, first
row). Finally, \gls*{t2b} confirmed that its correction
can not be compared, even when using the intensity compensation
based on the determinant of the deformation field.

@@

For the shake of completeness, we also document the outcome
of tractography at each step of the workflow. After distortion,
we obtained 1750 tracks and a characteristic length of 
$40.64\pm8.12$mm.
\gls*{fmb} produced 1900 tracks/$40.53\pm8.07$mm., \gls*{reb}
obtained 2200 tracks/$41.15\pm7.60$mm., and 2500 
tracks/$40.60\pm7.81$mm. for \gls*{t2b}. Although track counting has
been used as a measurement of success and as a weight of connectivity 
strength, the number of tracks produced by tractography strongly rely
on the seeding performed and the smoothing caused by interpolation used
within correction methodologies.

@@

Even though all the surveyed methods produced visually 
sound results, our study suggested that \gls*{reb} is the 
most accurate methodology for correction of susceptibility-induced
distortions in \gls*{dwi}. Moreover, \gls*{reb} also achieved
the best results in terms of connectivity, being able to recover
one of two lost connections after distortion. \Gls*{fmb} achieved 
very positive scores. Finally, the \gls*{t2b} method did not achieve
the necessary high-standards to recommend its use. We understand
that specific methods with anisotropic regularization that completely
restrict deformations to the phase-encoding direction would perform
significantly better compared to the standard method included
in this study.

One relevant limitation of this work is that the \gls*{fmb} is
used both in distortion creation and correction, what could
bias results. Nonetheless, \gls*{reb} performed better that \gls*{fmb}.
A second limitation is found on the impossibility of integrating the
point spread function-based methods for two reasons.
Firstly, the unmet need of a point spread function map consistent
with the synthetic phase-difference map. Secondly, to our knowledge,
there are no implementations of the method readily available.
In practice, this second limitation is narrowed down by the
fact that typical acquisition protocols usually include one or more
datasets among field mapping, reversed-encoding \textit{b0} or T2,
but they do not include the point-spread function mapping.

Future extensions of this work will evaluate
different realizations of the synthetic phase-difference map,
better characterizing the phenomenon. Moreover, the
framework can be enhanced for studying the impact of 
other artifacts as subject's motion or eddy 
currents-derived distortions.

### 2013-10-01 Publications/2013-ISBI14/05-conclusions.tex
Chus' changes

This paper proposes an evaluation framework to
comprehensively analyze the impact of susceptibility
induced distortion on tractography from \gls*{dmri}
data. Inaccuracy on tractography produce errors
and increased variability in subsequent connectivity 
analyses of the whole-brain. Specially, in regions 
influenced by  significant distortion.
We publicly release the framework and also contribute
with the evaluation of the most widely-used and
readily-available correction methodologies. 
This work shows that \gls*{dmri} should be corrected
for susceptibility-induced artifacts prior to tractography
and connectivity analyses. In that sense, the \gls*{reb} 
method was the best classified
in two of the three rankings (geometric 
correctness and signal recovery), whereas \gls*{fmb}
achieved the lowest impact on the connectivity matrix
extracted from tractography.

>>>>

This paper proposes an evaluation framework to
comprehensively analyze the impact of susceptibility
induced distortion on tractography from \gls*{dmri}
data. Inaccuracy on tractography produce errors
and increased variability in subsequent connectivity 
analyses of the whole-brain. Especially, in regions 
influenced by  significant distortion.
We publicly release the framework and also contribute
with the evaluation of the most widely-used and
readily-available correction methodologies. 
The reverse-encoding method (\gls*{reb}) achieved the
best results correcting for susceptibility-induced 
artifacts. We reported geometrical accuracy, signal
dropout recovery and connectivity matrix detection as
ranking indices.
